---
title: "Bias in removal models"
author: "Olivier Gimenez"
date: "Dec 2024"
output: 
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      dpi = 600, 
                      fig.height=6, 
                      fig.width = 1.777777*6)
library(tidyverse)
theme_set(theme_light())
```

# Objective

Compute the relative bias (in %) in the estimator of expected abundance $E(N) = \lambda$ in a removal model with constant parameters with package `unmarked` (frequentist/mle). Relative bias is $(E(\hat \lambda) - \lambda)/\lambda$.

Load packages:
```{r}
library(unmarked)
library(tidyverse)
```

# Function

Create function to carry out simulations:
```{r}
removal_par <- function(
    
  nb_sites = 10, # number of sites
  nb_occasions = 3, # number of occasions
  lambda = 50, # Poisson rate 
  p = 0.5, # capture probability
  n_sim = 10){ # number of simulations
  
  # preallocate memory for storing estimates
  res <- rep(NA, n_sim)
  
  # simulate data from a static occupancy model n_sim times
  for (j in 1:n_sim){
    
    # define state process
    N <- rpois(nb_sites, lambda) # abundance
    
    # pre-allocate memory for matrix of removal data
    y <- matrix(NA, nrow = nb_sites, ncol = nb_occasions) # removal counts
    
    # get multinomial cells probabilities
    pi <- rep(NA, nb_occasions)
    pi[1] <- p
    tmp <- rep(1 - p, nb_occasions - 1)
    for (k in 2:nb_occasions){
      pi[k] <- prod(tmp[1:(k-1)]) * p
    }
    
    # define observation process
    for(l in 1:nb_sites) {
      y[l,] <- rmultinom(1, N[l], c(pi, 1 - sum(pi)))[1:length(pi)]
    }
    
    # format data
    umf <- unmarkedFrameMPois(y = y, type = "removal")

    # fit removal model w/ constant parameters
    fm <- multinomPois(~1 ~1, umf, engine = "C")

    # get estimate of abundance
    tmp <- backTransform(fm, type = 'state')@estimate
    if (tmp > lambda * 2) next # probably a convergence issue
    res[j] <- tmp
    
    # return relative bias in %
    bias <- round(mean((res - lambda)/lambda, na.rm = TRUE) * 100,1)
    return(bias)
    
  } # end of simulations
} # end of function
```

# Bias calculation

Initialize a table for storing results:
```{r}
sim <- tibble(capture = double(), 
              abundance = double(),
              nsites = double(),
              noccasions = double(),
              bias = double())
```

Create a grids of values for the capture probability, abundance, the number of sites and the number of occasions:
```{r}
capture <- seq(0.3, 0.9, length = 20)
abundance <- round(seq(10, 100, length = 20))
nb_sites <- c(1, 5, 10, 50)
nb_occasions <- c(3, 5, 10)
```

Set the number of simulations to 500:
```{r}
n_sim <- 500
```

Compute bias:
```{r}
for (k in nb_sites){
  for (l in nb_occasions){
    for (i in capture){
      for (j in abundance){
        res <- removal_par(nb_sites = k,
                           nb_occasions = l,
                           lambda = j,
                           p = i,
                           n_sim = n_sim)
        sim <- sim %>% add_row(capture = i,
                               abundance = j,
                               nsites = k,
                               noccasions = l,
                               bias = res)
      }
    }
    print(paste0("Relative bias in abundance for ",k," sites and ",l," occasions: done"))
  }
}
```

Save results:
```{r}
saveRDS(object = sim, file = "sims/sims.rds")
```

# Visualization

Plot a heatmap of relative bias as a function of capture probability and abundance, for given values of the number of sites and occasions, and save the plot:
```{r}
# Create a custom labeller function
facet_labels <- function(variable) {
  paste0("Sites: ", variable$nsites, ", Occasions: ", variable$noccasions)
}

ggplot(sim, aes(x = capture, y = abundance, fill = bias)) +
  geom_tile() +  # Creates the heatmap
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  facet_wrap(
    vars(nsites, noccasions),
    ncol = 3,
    labeller = labeller(.multi_line = FALSE, 
      nsites = function(x) paste0("Sites: ", x),
      noccasions = function(x) paste0("Occasions: ", x)
    )
  ) +  
  labs(
#    title = "Relative bias in abundance",
    x = "Capture",
    y = "Abundance",
    fill = "Bias (%)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 11),  
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 12),
    axis.title.x =  element_text(size = 14),
    axis.title.y =  element_text(size = 14)
  )
```

Save plot:
```{r}
ggsave(filename = "sims/heatmap_bias.png")
```


