---
title: "Assessment of abundance estimator properties in removal models when the closure assumption is not met"
author: "Olivier Gimenez"
date: "Apr 2025"
output: 
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      dpi = 600, 
                      fig.height=6, 
                      fig.width = 1.777777*6)
library(tidyverse)
theme_set(theme_light())
```

# Objective

N-mixture models assume *population closure*; that is, $N$ remains constant over all occasions; there is no mortality, recruitment or movement into and out of the population during the time between the first and last occasions. What if this assumption is not met? 

Here I simulate data from an open-population model to break that closure assumption. Specifically, I generate data where individuals can leave (emigration) and/or arrive between occasions (immigration). We may think of these two terms also as survival and recruitment.

To assess performance, I compute the relative bias (in %) in the estimator of expected abundance $E(N) = \lambda$ in a removal model with constant parameters with package `unmarked` (frequentist/mle). Relative bias is $(E(\hat \lambda) - \lambda)/\lambda$. I also report the root mean square error and the coverage of the confidence interval. 
 
Load packages:
```{r}
library(unmarked)
library(tidyverse)
```

# Functions

Create functions to carry out simulations:
```{r}
# simulate data that break the closure assumption
sim_open <- function(nb_sites, nb_occasions, lambda, p, stay, arrivals) {
  # nb_sites: number of sites
  # nb_occasions: number of occasions
  # lambda: expected abundance, Poisson rate 
  # p: capture probability
  # stay: Pr(individual sticks around between occasions)
  # arrivals: mean number of new arrivals between occasions
  y <- matrix(0, nb_sites, nb_occasions)
  for(i in 1:nb_sites) {
    avail <- rpois(1, lambda) # population at first sampling occasion
    for(j in 1:nb_occasions) {
      y[i, j] <- rbinom(1, avail, p) # detect and remove
      avail <- avail - y[i, j] # left after removal
      if(j < nb_occasions) { # between‑occasion dynamics
        avail <- rbinom(1, avail, stay) +  # survivors who stick around
          rpois(1, arrivals)               # new arrivals
      }
    }
  }
  y
}

# simulate, fit and compute metrics
removal_par <- function(
  nb_sites = 5, # number of sites
  nb_occasions = 3, # number of occasions
  stay = 0.80, # Pr(individual sticks around between occasions)
  arrivals = 1.0, # mean number of new arrivals between occasions
  lambda = 50, # Poisson rate 
  p = 0.5, # capture probability
  n_sim = 10){ # number of simulations
  
  # preallocate memory for storing estimates
  res <- rep(NA, n_sim)
  low <- rep(NA, n_sim)
  up <- rep(NA, n_sim)
  
  # simulate data n_sim times
  for (j in 1:n_sim){

    # simulate data that break the closure assumption
    y <- sim_open(nb_sites = nb_sites, 
                  nb_occasions = nb_occasions, 
                  lambda = lambda, 
                  p = p, 
                  stay = stay, 
                  arrivals = arrivals)
      
    # format data
    umf <- unmarkedFrameMPois(y = y, type = "removal")

    # fit removal model w/ constant parameters
    fm <- multinomPois(~1 ~1, umf, engine = "C")

    # get estimate of abundance
    zz <- backTransform(fm, type = 'state')
    tmp <- zz@estimate
    if (tmp > lambda * 2) next # probably a convergence issue
    res[j] <- tmp
    # get 95% confidence interval
    ci <- confint(zz, level = 0.95)
    low[j] <- ci[1]
    up[j] <- ci[2]
  } # end of simulations
  
  # compute relative bias in %
  bias <- round(mean((res - lambda)/lambda, na.rm = TRUE) * 100,1)
  # compute root mean square error
  rmse <- round(sqrt(mean((res - lambda)^2, na.rm = TRUE)), 3)
  # compute coverage of 95% confidence interval
  cover <- mean(lambda >= low & lambda <= up, na.rm = TRUE)
  out <- list(bias = bias,
              rmse = rmse,
              cover = cover)
  return(out)
} # end of function
```

# Metrics calculation

Initialize a table for storing results:
```{r}
sim <- tibble(capture = double(), 
              abundance = double(),
              nsites = double(),
              noccasions = double(),
              bias = double(),
              rmse = double(),
              cover = double())
```

Create a grids of values for the capture probability, abundance, the number of sites and the number of occasions:
```{r}
capture <- seq(0.3, 0.9, length = 20)
abundance <- round(seq(10, 100, length = 20))
nb_sites <- c(1, 5, 10, 50)
nb_occasions <- c(3, 5, 10)
```

Set the number of simulations to 100:
```{r}
n_sim <- 100
```

Compute metrics:
```{r}
for (k in nb_sites){
  for (l in nb_occasions){
    for (i in capture){
      for (j in abundance){
        res <- removal_par(nb_sites = k,
                           nb_occasions = l,
                           stay = 0.80,
                           arrivals = 1.0, 
                           lambda = j,
                           p = i,
                           n_sim = n_sim)
        sim <- sim %>% add_row(capture = i,
                               abundance = j,
                               nsites = k,
                               noccasions = l,
                               bias = res$bias,
                               rmse = res$rmse,
                               cover = res$cover)
      }
    }
    print(paste0("Calculations for ",k," sites and ",l," occasions: done"))
  }
}
```

Save results:
```{r, eval=FALSE}
saveRDS(object = sim, file = "sims/simsclosure100.rds")
```


# Visualization

## Relative bias

Plot a heatmap of relative bias as a function of capture probability and abundance, for given values of the number of sites and occasions, and save the plot:

```{r}
# Create a custom labeller function
facet_labels <- function(variable) {
  paste0("Sites: ", variable$nsites, ", Occasions: ", variable$noccasions)
}

ggplot(sim, aes(x = capture, y = abundance, fill = bias)) +
  geom_tile() +  # Creates the heatmap
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  facet_wrap(
    vars(nsites, noccasions),
    ncol = 3,
    labeller = labeller(.multi_line = FALSE, 
      nsites = function(x) paste0("Sites: ", x),
      noccasions = function(x) paste0("Occasions: ", x)
    )
  ) +  
  labs(
    x = "Capture",
    y = "Abundance",
    fill = "Bias (%)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 11),  
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 12),
    axis.title.x =  element_text(size = 14),
    axis.title.y =  element_text(size = 14)
  )
```

Save plot:
```{r}
ggsave(filename = "sims/heatmap_bias_closure.png")
```


## Root mean square error

Plot a heatmap of rmse as a function of capture probability and abundance, for given values of the number of sites and occasions, and save the plot:
```{r}
# Create a custom labeller function
facet_labels <- function(variable) {
  paste0("Sites: ", variable$nsites, ", Occasions: ", variable$noccasions)
}

ggplot(sim, aes(x = capture, y = abundance, fill = rmse)) +
  geom_tile() +  # Creates the heatmap
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  facet_wrap(
    vars(nsites, noccasions),
    ncol = 3,
    labeller = labeller(.multi_line = FALSE, 
      nsites = function(x) paste0("Sites: ", x),
      noccasions = function(x) paste0("Occasions: ", x)
    )
  ) +  
  labs(
#    title = "Relative bias in abundance",
    x = "Capture",
    y = "Abundance",
    fill = "RMSE"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 11),  
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 12),
    axis.title.x =  element_text(size = 14),
    axis.title.y =  element_text(size = 14)
  )
```

Save plot:
```{r}
ggsave(filename = "sims/heatmap_rmse_closure.png")
```


## Coverage

Plot a heatmap of the coverage of 95% confidence interval as a function of capture probability and abundance, for given values of the number of sites and occasions, and save the plot:
```{r}
# Create a custom labeller function
facet_labels <- function(variable) {
  paste0("Sites: ", variable$nsites, ", Occasions: ", variable$noccasions)
}

ggplot(sim, aes(x = capture, y = abundance, fill = cover)) +
  geom_tile() +  # Creates the heatmap
  viridis::scale_fill_viridis(
    option = "plasma",  
    direction = 1,     # rouge → jaune → violet (inverse si besoin)
    limits = c(0, 1)
  ) + 
#  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = ) +
  facet_wrap(
    vars(nsites, noccasions),
    ncol = 3,
    labeller = labeller(.multi_line = FALSE, 
      nsites = function(x) paste0("Sites: ", x),
      noccasions = function(x) paste0("Occasions: ", x)
    )
  ) +  
  labs(
#    title = "Relative bias in abundance",
    x = "Capture",
    y = "Abundance",
    fill = "Coverage"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 11),  
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 12),
    axis.title.x =  element_text(size = 14),
    axis.title.y =  element_text(size = 14)
  )
```

Save plot:
```{r}
ggsave(filename = "sims/heatmap_cover_closure.png")
```
